---
title: "Visualizations - STARTER"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
---

```{r message = FALSE, warning = FALSE}

# load packages
library(tidyverse)

```

## Example dataset 

```{r}

```

### Aesthetic mapping 

```{r}

ggplot(data = diamonds)

```

### Exercise 

- Use the code chunk below to create a scatterplot for `table` vs `depth`.

```{r}

```

### Other attributes 

```{r}

ggplot(data = diamonds,
       aes(x = carat,
           y = price)) + 
  geom_point()

```

```{r error = TRUE}

ggplot(data = diamonds,
       aes(x = carat,
           y = price)) + 
  geom_point(color = purple)

```

### Incorporating more variables via `aes()` 

```{r}

ggplot(data = iris,
       aes(x = Sepal.Length,
           y = Sepal.Width))

```

### Exercise 

- Using the code chunk below, create a scatterplot using a sample of the _diamonds_ dataset that includes the following features (code for the sample is provided, check the help page to see how it works!):

  a) Visualizes `table` vs `depth`;
  
  b) Each observation is sized based on the weight `carat`;
  
  c) All observations have the shape of a diamond (*HINT: Google "geom_point shapes"*).

```{r}

diamonds_sample <- sample_n(diamonds, size = 100)

```

## Histograms 

### geom_histogram() 

```{r}

ggplot(data = diamonds)

```


```{r}

ggplot(data = iris,
       aes(x = Petal.Length)) + 
  geom_histogram()

```

### Exercise 

- Use the code chunk below to create a histogram for `carat` that has 20 bins with white bars and black outlines.

```{r}

```

### Titles and labels 

```{r}

ggplot(data = diamonds,
       aes(x = carat)) + 
  geom_histogram(binwidth = 0.1,
                 fill = "red",
                 color = "purple")

```

### Exercise 

- Using the code chunk below, modify the code used to make the previous plot to:

  a) Visualize sepal length instead of petal length;
  
  b) Change the X-axis label to be more readable;
  
  c) Give the plot an informative title.

```{r}

```

## Boxplots 

### geom_boxplot() 

```{r}

x <- rnorm(50)
summary(x)

```

```{r}

ggplot(data = diamonds)

```

### Exercise 

- Using the code chunk below, create a *vertical* boxplot of `Sepal.Length` using the _iris_ dataset.

```{r}

```

### Comparitive boxplots 

```{r}

ggplot(data = diamonds,
aes(x = carat)) + 
  geom_boxplot()

```

### Transformations and more customizations 

```{r}

ggplot(data = diamonds,
       aes(x = cut,
           y = price)) + 
  geom_boxplot(fill = "darkblue",
               color = "lightblue")

```

### Exercise 

- Using the code chunk below, create a comparative boxplot using the _iris_ dataset that includes the following features:

  a) Compares `Sepal.Length` for each `Species`;
  
  b) Horizontal boxplots (more than one way to do this);
  
  c) `Sepal.Length` is originally measured in millimeters (mm); convert this to meters (m) for this plot;

  d) More informative axis label for `Sepal.Length` based on new scale;
  
  e) A cool theme (try a few!).

```{r}

```

## Bar graphs 

### Bar graphs

### geom_bar() wtih raw data 

```{r]}

ggplot(data = diamonds,
       aes(x = cut))

```

### geom_bar() wtih count data 

```{r}

table(cut = diamonds$cut)
cut_table <- count(diamonds, cut)
glimpse(cut_table)

```

```{r}

ggplot(data = diamonds,
       aes(x = cut)) + 
  geom_bar()

```

```{r}

ggplot(data = cut_table,
       aes(x = cut,
           y = n))

```

### Exercise 

- Using the code chunk below, create a bar graph of `Species` using a sample from the _iris_ dataset provided below. *HINT: Make sure to inspect the data first*.

```{r}

# create a sample from of data and summarize
# -> rename so in slightly different format than example and have to work with it
iris_sample <- iris %>% 
  sample_n(size = 50) %>% 
  count(Species) %>% 
  rename(Count = n)

```

```{r}

iris_sample

```

### Stacked bar graph 

```{r}

ggplot(data = diamonds,
       aes(x = cut)) + 
  geom_bar()

```

### Proportionally stacked bar graph 

```{r}

ggplot(data = diamonds,
       aes(x = cut,
           fill = clarity)) + 
  geom_bar()

```

### Side-by-side bar graph 

```{r}

ggplot(data = diamonds,
       aes(x = cut,
           fill = clarity)) + 
  geom_bar(position = "fill")

```

```{r}

```

## Line plots 

### geom_line()

```{r}

data_sun <- data.frame(year = c(1700:1988),
                       sunspots = as.vector(sunspot.year))
glimpse(data_sun)

```

```{r}

ggplot(data = data_sun)

```

## Facets 

### Facets 

### facet_wrap() 

```{r}

ggplot(data = diamonds,
       aes(x = table,
           y = depth)) + 
  geom_point()

```

### Controlling rows and columns 

```{r}

ggplot(data = diamonds,
       aes(x = carat)) + 
  geom_boxplot() + 
  facet_wrap(~ cut)

```

```{r}

ggplot(data = diamonds,
       aes(x = price)) + 
  geom_histogram() + 
  facet_wrap(~ clarity,
             nrow = 1)

```

### Scales for facets 

```{r}

ggplot(data = diamonds,
       aes(x = price)) + 
  geom_histogram() + 
  facet_wrap(~ cut)

```

### Exercise 

- Using the code chunk below, create a boxplots using the _iris_ dataset that includes the following features:

  a) Plots `Sepal.Length` faceted by `Species`;
  
  b) Stack the facets vertically to emphasize the differences between each Species;
  
  c) Think about if we want to adjust the scales of the panel, why or why not?
  
- Note that we made this plot previously with comparitive boxplots by specifying `aes(y = Species)`; so this is an alternative.

```{r}

```

### facet_grid() 

```{r}

ggplot(data = diamonds,
       aes(x = table)) + 
  geom_boxplot()

```

### Exercise 

- Using the code chunk below, create scatterplots of `carat` vs `price` that is colored by `cut` and facetted by `color` and `clarity`. Try a few different options for `scales` to see how they affect the plot.

```{r}

# create a sample from of data with fewer levels (variety from ordinal scale of each)
diamonds_sample <- diamonds %>% 
  filter(cut %in% unique(diamonds$cut)[c(1,3,5)],
         color %in% unique(diamonds$color)[c(1,4,7)],
         clarity %in% unique(diamonds$clarity)[c(1,4,8)]) %>% 
  sample_n(size = 5000)

```

```{r}

diamonds_sample

```

## Density histograms and density curves 

### Density histograms 

```{r}

ggplot(data = diamonds,
       aes(x = carat)) + 
  geom_histogram() + 
  facet_wrap(~ cut,
             scales = "free_y")

```

### Density curves 

```{r}

ggplot(data = diamonds,
       aes(x = carat,
           y = after_stat(density))) + 
  geom_histogram(bins = 10)

```

```{r}

ggplot(data = diamonds,
       aes(x = carat))

```

### Exercise 

```{r}

```

### Multiple density curves 

```{r]}

ggplot(data = diamonds,
       aes(x = carat)) + 
  geom_density()

```

### Amount of smoothing 

```{r}

ggplot(data = diamonds,
       aes(x = carat)) + 
  geom_density()

```

### Exercise 

- Using the code chunk below, use the _iris_ dataset to do the following:

  a. Create density curves of a `Petal.Length` colored by each `Species`;
  
  b. Adjust the amount of smoothing to find a level that shows the overall trends well with some detail, but that is not overly exact (it's a balance);
  
  c. Add `fill = Species` in the `aes()` statement to see the result of this;
  
  d. Notice how the colors are completely opaque (not "see through"). To make them more transparent, add the following option locally `geom_density(alpha = 0.5)`. Try a few different values (`0 \le alpha \le 1`) to see the result. Note that this option can also be used with lots other `geom_*()` as well.
  
  e. What does this plot tell us about the petal lengths of the different species?

```{r}

```

## Application 

### Recreating 

```{r eval = FALSE}

# load basketball data
# TO DO FIRST:
# -> 1) create a folder named "data" in the same location where you saved this file and save the file "data-bsu-game.RData" in it
# -> 2) remove eval = FALSE from the code chunk header
load("data/data-bsu-game.RData")
data_bsu_game

```

### Exercise 

- Using the code chuck below, copy the final code used to create `Visual 1' from above and modify it to recreate 'Visual 2'.

```{r}


```