---
title: "Reading Check Quizzes - Storytelling With Data"
format: html
---

### Instructions and Canvas 

Answer the following questions after ***reading pages # - #*** of ***Storytelling With Data***.

NOTE: Canvas can shuffle questions (and answers if want).

- Putting the correct answer first always here and can reorder when inputting into Canvas.

- If there is not an "All of the above" type answer, then will shuffle answers.

### Grading 

Overall grading

- Each quiz counts for the same percent of overall grade

Points assignment

- All quizzes will have 3 or 4 questions

- Equally weighted point totals for each question (with partial credit for select all that apply / multiple answer questions)

- With equal weighting from above, point totals don't matter (so leaving as default pts = 1).

### Reading Check Quiz 1 

**SWD: Pages 1 --- 13.5** (up to "How this book us organized" section)

Introduction (Storytelling; How you will learn with this book)

*OPTIONAL: Read the Foreword and **About the Author sections if interested!*

Note that you only get *1 attempt (no time limit)*.

1.  Which of the following are reasons why visualizations are often *ineffective*? (Select all that apply)

    a.  **We were never taught how to tell stories with numbers.**

    b.  **Relying on tool defaults.**

    c.  **There are gaps in capabilities for those that are creating the visuals.**

    d.  **Ending the visualization with the first graph we make.**
    
2.  Typically, which part of the analysis process does the *audience* see? (Multiple Choice)

    a.  **Communication of the analysis**
    
    b.  Finding the data

    c.  Analyzing the data

    d.  Building models
    
3.  Which software(s) to the *lessons* of Storytelling With Data *apply* to? (Multiple Choice)

    a.  Excel

    b.  Tableau

    c.  R

    d.  **All of the above**

### Reading Check Quiz 2 

**SWD: Pages 13.5 --- 27.5** (up to "Who, what, and how: illustrated by example" section)

Introduction (Organization of book)

Ch 1 Importance of Context (Types of analyses; Questions to ask; Mechanism)

Note that you only get *1 attempt (no time limit)*.

1.  When *planning* an explanatory analysis, what is the correct *order* of questions to answer? (Drag into order, on Canvas match to first, second, third)

    a.  **To whom are you communicating?**

    b.  **What do you want your audience to know?**

    c.  **How can you use data to help make your point?**

2.  Which of the following statements are *true* regarding how to *create* your presentations? (Select all that apply)

    a.  **Slides should be sparse**
    
    b.  **You should be able to answer any audience questions regardless of whether that detail is in the presentation or not**

    c.  Slides should act as your speaker notes

    d.  Slides should be detailed enough to answer any potential questions the audience may have
    
3.  Which of the following aspects are impacted by the *method* of the communications? (Select all that apply)

    a.  **The amount of control you will have over how the information is consumed**

    b.  **The level of detail**

    c.  The analysis that is performed

    d.  The effectiveness of the communication
    
### Reading Check Quiz 3 

**SWD: Pages 27.5 --- 33** (up to end of Chapter 1)

Ch 1 Importance of Context (Who, what, how example; 3-min story; Big idea; Storyboarding)

Note that you only get *1 attempt (no time limit)*.

1.  Which of the following are *not benefits* of creating a *3-minute story*? (Multiple Choice)

    a.  **Can use the 3-minute story directly in your presentation.**

    b.  Removes dependence on your slides or visuals for a presentation.

    c.  As a consequence, you know what the most important pieces are and what isn't essential.
    
    d.  Ensures you are clear on and can articulate the story you want to tell.

2.  What is the recommended *way* to *storyboard*? (Multiple Choice)

    a.  **Use a whiteboard, post-it notes or plain paper.**
    
    b.  Create slides and choose the most important ones.

    c.  Write the report and get key ideas from each section.

    d.  Convert your analysis to slides using software.
    
3.  Which of the following are the *advantages* of *storyboarding*? (Multiple choice)

    a.  Avoid attachment to slides that were made before forming a narrative.

    b.  Can easily rearrange, add, or remove pieces to explore different narrative flows.

    c.  Saves time by reducing the number of iterations when building content.

    d.  **All of the above.**
    
### Reading Check Quiz 4 

**SWD: Pages 35 --- 55.5** (up to "Waterfall chart" section)

Ch 2 Choosing an effective visual (Simple text; Tables; Heatmap; Graphs; Points; Lines; Bars)

Note that you only get *1 attempt (no time limit)*.

1.  Which of the following statements is *not true about tables*? (Multiple choice)

    a.  **Audiences interact with tables and graphs the same way.**

    b.  If used in a presentation, you can lose your audience's attention as they read the table.

    c.  The design of the table needs to be considered, specifically how to let the data be the focus.
    
    d.  Heatmaps add visual cues to a table, which can make points of interest, such as low and high values, stand out more.
    
    e.  Tables are great for communicating to a mixed audience who will search for their particular row of interest.

2.  Match each *graph* to the *type of data* it displays. (Matching)

    a.  Scatterplot: Relationship between two variables.
    
    b.  Line graph: Continuous data, often over time.

    c.  Slopegraph: Comparison between two points across categories.

    d.  Bar graph: Categorical data (organized in groups) 
    
3.  Which of the following statements are *true about bar graphs*? (Select all that apply)

    a.  **Because of how our eyes compare the relative end points of the bars, it is importaant that bar graphs always have a zero baseline.**

    b.  **In general, bars should be wider than the white space between them, but not too wide or the audience will start to compare areas instead of lengths.**

    c.  **There are multiple types of bar graphs that can solve different visualization problems.**

    d.  Bar graphs should be avoided because they are common.
    
    e.  All subcomponent categories in a stacked vertical bar charts are easy to compare with appropriately chosen colors.
    
### Reading Check Quiz 5 

**SWD: Pages 55.5 --- 69** (up to end of Chapter 2)

Ch 2 Choosing an effective visual (Bars; Area; Other types of graphs; To be avoided)

Note that you only get *1 attempt (no time limit)*.

1.  How do we typically process information on a chart? What path do our eyes make across the screen or page? (Multiple choice)

    a.  **Letter Z path.**

    b.  Left to right.

    c.  Top to bottom.
    
    d.  Random, no pattern.

2.  All of the following graphs should be avoided. Match each *graph* to the *correct reason* for *why it should be avoided*. (Matching)

    a.  Pie chart: The human eye isn't good at interpreting areas (ascribing quantitative information to two-dimensional space).
    
    b.  Donut chart: Arc lengths are even harder than areas to compare.

    c.  3D: Introduces unnecessary chart elements and visual distortions.

    d.  Secondary y-axes: Takes time and reading to understand how to read the data.
    
3.  Instead of a *secondary y-axis*, there are a few alternatives. Match each *strategy* with its *advantage.* (Matching)

    a.  Label directly: Puts more attention on the specific numbers.
    
    b.  Pull apart vertically: Puts more emphasis on the overarching trends.

    c.  Color data to match axis: This can work, but this element can often be used more strategically.
    
### Reading Check Quiz 6 

**SWD: Pages 71 --- 84.5** (up to "White space" section)

Ch 3 Clutter is your enemy (Cognitive load; Clutter; Gestalt principles of visual perception; Lack of visual order, alignment)

Note that you only get *1 attempt (no time limit)*.

1.  Which of the following are reasons to *reduce clutter*? (Select all that apply)

    a.  **To reduce the perceived cognitive load on the audience.**

    b.  **Clutter makes our visuals appear more complicated than necessary.**

    c.  **If cluttered, we risk the audience deciding they do not want to take the time to understand what we are showing them.**

    d.  Clutter is needed to fill white space to show more information.
    
2.  Which *Gestalt principle of visualization* is leverage by *shading the area behind forecasts* in time series plots? (Multiple Choice)

    a.  **Enclosure**
    
    b.  Proximity

    c.  Similarity

    d.  Closure
    
    e.  Contrast
    
3.  Which of the following is *true* about *text on plots*? (Select all that apply)

    a.  **The goal is to create clean lines both horizontally and vertically with alignment of elements and white space.**
    
    b.  **Text such as titles, subtitles, axis titles, and legends should be upper-left most justified.**

    c.  Diagonal components are okay when there isn't enough space.
    
    d.  Should always draw lines to connect text blocks to parts of the plot they refer to.
    
### Reading Check Quiz 7 

**SWD: Pages 84.5 --- 98** (up to end of Chapter 3)

Ch 3 Clutter is your enemy (Lack of visual order, white space; Non-strategic use of contrast; De-cluttering)

**R4DS: Chapter 3 -- Workflow basics**

Coding basics, comments, what's in a name, calling functions

Note that you only get *1 attempt (no time limit)*.

1.  Which of the following are *true* about the *minimal guidelines* when it comes to *preserving white space*? (Select all that apply)

    a.  **Margins should remain free of text and visuals.**

    b.  **Size visuals appropriately to their content, not to fill up space.**

    c.  **White space should be used strategicially.**

    d.   White space should be filled.
    
    e.   White space isn't an important consideration in design.
    
2.  We should use *data markers / labels*...? (Multiple Choice)

    a.  **On purpose and with a purpose.**
    
    b.  Always.

    c.  Becuase they are the default in applications.

    d.  If they look nicer.
    
3.  *Labeling data directly* and with the *same color* takes *advantage* of which *two Gestalt principles of visualization*? (Multiple Choice)

    a.  **Proximity and similarity**
    
    b.  Proximity and closure

    c.  Similarity and color
    
    d.  Connection and enclosure
    
4. Which of the following *comments* has the correct *content* according to the author. (Multiple choice)

  a. # Manually change the colors of the points to bring emphasis to the specific points of interest and make the others fade into background
  
  b. # Manually change the colors using scale_color_manual(), have to match each value to each color
  
  c. # Create a plot with different colored points
    
### Reading Check Quiz 8 

**SWD: Pages 99 --- 116.5** (up to "Size" section)

Ch 4 Focus your audience's attention (Memory; Preattentive attributes, signal where to look in text and in graphs)

**R4DS: Chapter 5 -- Code style**

Names, spaces, pipes, ggplot2, sectioning comments

Note that you only get *1 attempt (no time limit)*.

1.  Which of the following are *true* about the *preattentive attributes*? (Select all that apply)

    a.  **Iconic (super fast) memory is tuned to preatttentive attributes, which enables our audience to see what we want them to see before they know they are seeing it.**

    b.  **Preattentive attributes change the way you process information.**

    c.  **Preattentive attributes can be leveraged in text and graphs.**

    d.  Preattentive attributes are for visualizations only.
    
    e.  All preattentive attributes have equal strength in grabbing the audience's attention.
    
2.  Which of the following are statements are *true* regarding the use of *repeated iterations of the same plot* in a *presentation setting*? (Select all that apply)

    a.  **This strategy can familiarize your audience with the data and visual first before highlighting the important aspects.**
    
    b.  **You can tell different stories by emphasizing different aspects.**

    c.  The benefits are limited because you can only leverage one to two preattentive attributes for a plot.

    d.  The main benefit is ease of prep work for the presenter.
    
3.  Which of the following statements are *true* for *how you should start when creating plots*? (Select all that apply)

    a.  **A good strategy is to push everything to the background.**
    
    b.  **When creating exploratory plots at the start, you should avoid the use of preattentive attributes.**

    c.  **By pushing everything to the background, you have to make explicit decisions about what to bring to the forefront or highlight.**
    
    d.  For exploratory analysis, the first plots should already have a plan for what preattentive attributes to leverage.
    
4. Which of the following *code chunks* has the *recommended style* according to the author (ignoring the difference between base R pipe |> and %>%)? (Multiple choice)

  a.
  
```{r, eval = FALSE}
rnorm(n = 10) %>% 
  scale %>% 
  mean(rm.na = TRUE)
```
  
  b.
  
```{r, eval = FALSE}
rnorm(n=10) %>% scale %>% mean(na.rm=TRUE)
```
  
  c.
  
```{r, eval = FALSE}
mean(scale(rnorm(n = 10)), na.rm = TRUE)
```
  
### Reading Check Quiz 9 

**SWD: Pages 116.5 --- 126** (up to end of Chapter 4)

Ch 4 Focus your audience's attention (Size; Color; Position on page)

Note that you only get *1 attempt (no time limit)*.

1.  Which of the following are *true* about the use of *color*? (Select all that apply)

    a.  **The authors suggest shades of grey and a single bold color to draw attention such as blue or black.**

    b.  **Color should be used sparingly, consistently, and designed with the color-blind in mind.**

    c.  **Color saturation (think intensity) can cary quantitative information.**

    d.  Color should be used to to make the plots not lose the audience's attention.
    
    e.  Color does not envoke emotion or have connotations associated with them.
    
2.  Which pair of colors does the author suggest as *substitutes* for *green / red* for *positive / negative* values, respectively? (Multiple choice)

    a.  **Blue / orange**
    
    b.  Purple / yellow

    c.  Grey / orange

    d.  Blue / yellow
    
3.  Where on the page should *important things* go for *plots and slides*? (Multiple choice)

    a.  **Top left**
    
    b.  Top right

    c.  Bottom left
    
    d.  Bottom right
    
    e.  Wherever fits best
    
### Reading Check Quiz 10 

**SWD: Pages 127 --- 138.5** (up to "Accessibility" section)

Ch 5 Think like a designer (Affordances (highlight the important stuff, eliminate distractions, visual hierarchy))

Note that you only get *1 attempt (no time limit)*.

1.  Which of the following are *true* about *visual affordances**? (Select all that apply)

    a.  **Visual affordances can be leveraged to indicate to our audience how to tuse and interact with our visualization.**

    b.  **Preattentive attributes can be used to highlight the important stuff, and be also can be layered.**

    c.  **Bold, uppercase, size and color are the most effective ways to highlight text while adding minimal noise.**

    d.  **At most 10% of the visual design should be highlighted.**
    
    e.  At most 20% of the visual design should be highlighted.
    
    f.  Italics and underline add the same amount of noise and can be used interchangeably.
    
2.  Which of the following are *true* about *distractions*? (Select all that apply)

    a.  **To identify distractions, we need to think about clutter and context.**
    
    b.  **To recognize distractions, we can recognize that not all data are equally important and some details can be summarized.**

    c.  **Each step in reduction and de-emphasis causes what remains to stand out more.**

    d.  If a detail is needed for the appendix, it should be included on the slide where it is mentioned because it is important enough.
    
    e.  To identify distractions, we should ask ourselves if anything else needs to be added to the visual?
    
3.  Which of the following are *true* about *visual hierarchy*? (Select all that apply)

    a.  **Design can influence th e order in which the audience interacts with a visual.**
    
    b.  **Super categories can be used in tables or plots (such as labeling quadrants on a scatterplot) to establish a hierarchy.**

    c.  All categories must always be labeled when utilizing super categories on a plot.
    
    d.  Audiences will always process the hierarchy in the same order.
    
### Reading Check Quiz 11 

**SWD: Pages 138.5 --- 150** (up to end of Chapter 5)

Ch 5 Think like a designer (Accessibility (don't overcomplicate, text); Aesthetics; Accepatance)

Note that you only get *1 attempt (no time limit)*.

1.  Which of the following are *true* about *accessibility**? (Select all that apply)

    a.  **Applied to data visualization, accessibility means plots should be designed to be easy to use by people of widely varying technical skills.**

    b.  **Poorly designed plots lead to misinterpretation and lack of understanding, which is the designers fault.**

    c.  **Every chart needs a title and every axis needs a title (exceptions are extrememly rare).**

    d.  Audienecs will always draw the same conlcusions from a well designed plot.
    
    e.  More complicated graphs mean that the conclusions that can be drawn are more advanced / technical.
    
    
2.  Which of the following are *advantages* of *pretty plots*? (Select all that apply)

    a.  **Pretty plots are perceived as easier to use.**
    
    b.  **Pretty plots are more accepted over time, promote creative thinking, and audiences are more tolerant of design flaws.**

    c.  Extra fluff should be added to make designs pretty.

    d.  One needs to study art and advanced design in order to be able to create aesthetic designs in their plots.
    
    
3.  When it comes to *aesthetics*, which of the following *lessons* need to be considered?  (Select all that apply)

    a.  **Be smart with color.**
    
    b.  **Pay attention to alignment.**

    c.  **Leverage white space.**
    
    d.  Add extra elements for more engaging plots.
    
    e.  Fill entire space.
    
    f.  Adds colors to be more aesthetically pleasing.